Policy Gradient Estimation

### 2019-03-06 pge-mattis.tex
Update LR derivation (still in progress)

\paragraph{Definitions.} 
We define states $s \in \mathbb{S}$, actions $a \in \mathbb{A}$, and rewards $r \in \mathbb{R}$. 
For discrete problems, we assume $\mathbb{S} \equiv \{0,1,2,\ldots\}$ and $\mathbb{A} \equiv \{0,1,2,\ldots\}$ without loss of generality.
A trajectory $\tau_{\pi_\theta}^h := (s_0, a_0, $ $s_1, a_1, \dots, s_h, a_h)$ is generated by drawing $s_0 \sim p(x_0)$ according to the distribution over initial states $p(x_0)$, and successively sampling $a_t \sim \pi_\theta(a_t|s_t)$ according to the policy $\pi_\theta$, and $s_{t+1} \sim p(s_{t+1}|s_t,a_t)$ until the horizon $h$ is reached. 
At each time step, we receive a reward according to $r_t = r(s_t, a_t, s_{t+1})$. 
A trajectory can also be called \textit{roll-out} or \textit{episode}, though the term episode implies it ends in a terminal state.

@@

\paragraph{Likelihood-ratio gradients.} 
For this derivation, we will change the perspective a bit requiring some additional definitions. 
We define ${\mu_\pi}_i = \sum_{t=0}^{\infty} p(s_t=i | s_0, \pi)$ \cite{puterman2014markov} as the state distribution, though it does not sum up to one without normalization. Note that $\mu_\pi$ is equivalent to the discounted state visit count $d^\pi$ introduced by Sutton et al. \cite{Sutton:1999:PGM:3009657.3009806}.
We define $P_\pi$ as the transition matrix, i.e. ${P_\pi}_{i,j}=p(s_j|s_i,\pi)$.
$r_\pi$ contains the mean rewards for all states. 
$\mu_0$ is the initial state distribution.
$\psi^\pi$ denotes some score function with $\psi_\pi(\tau) \propto \mathcal{R}^\tau$, see section \ref{sec:pge}, 
From this, we derive
\begin{align}
	\mu_\pi &= \mu_0 + P_\pi^T \mu_\pi \nonumber \\
	(I-P^T)\mu_\pi &= \mu_0 \nonumber \\
	\mu_\pi^T &= \mu_u (I-P_\pi)^{-1} ,
\end{align}
so we can reformulate the problem as 
\begin{align}
	&J(\theta) = \mu_0^T \psi^\pi \\
	\text{s.t.}\ &\psi^\pi = r^\pi = \gamma P^\pi \psi^\pi . \nonumber
\end{align}
From which it follows, that
\begin{align*}
	\nabla_\theta J(\theta) &= \mu_0^T \nabla_\theta \psi_\pi
	\\ &= \mu_0^T (I - \gamma P_\pi)^{-1} (\nabla_\theta r + \gamma (\nabla_\theta P_\pi) \psi_\pi)
	\\ &= \mu_\pi^T (\nabla_\theta r_\pi + \gamma (\nabla_\theta P_\pi) \psi_\pi)
	\\ &= \sum \mu_\theta(s) \nabla_\theta\pi(a|s) Q(s,a)
	\\ &= \sum \mu_\theta(s) \pi(a|s) \nabla_\theta\log\pi(a|s) Q(s,a)
\end{align*}

$r_i^\theta = \int_\mathbb{A} r(s,a)\pi_\theta(a|s) da$

Where we used $\nabla_\theta \pi_\theta(a|s) = \pi_\theta(a|s)\nabla_\theta\log\pi_\theta(a|s)$, obtained from the likelihood ratio, which gives us the likelihood-ratio gradient,
\begin{equation}
  \nabla_\theta \mathbb{E}_{\pi_\theta}\left[\psi(\tau)\right] = \mathbb{E}_{\pi_\theta}\left[\nabla_\theta{\log\pi_\theta(a|s)}\psi(\tau)\right].
  \label{eqn:like-grad}
\end{equation}
intuitively meaning we should to increase the probability of actions judged as good by the score function $\psi$. We present some frequently used score functions in section \ref{sec:pge}.

In this section, we discuss some approaches to calculate the estimated policy gradient $\nabla_\theta J(\theta) = \mathbb{E}_{\pi_\theta}\left[\nabla_\theta{\log\pi_\theta(a|s)}\psi(a|s)\right]$ using different score functions to substitute $\psi$.

\paragraph{Immediate updates.} 
The first class of algorithms developed to update a policy directly instead of deriving it from a value function approximation were REINFORCE \cite{Williams92simplestatistical} algorithms.
REINFORCE uses immediate updates with a baseline $b$ in its gradient steps:
\begin{equation}
  \psi_t \leftarrow r_t-b.
\end{equation}

Obviously, this does not involve temporal information. 
To credit temporal information $\psi_t$ needs to involve multiple time steps, meaning we need to delay updates until we have enough information to evaluate the gradient.

\paragraph{Episodic updates.} 
A common way of crediting temporal information in score function is to sample trajectories, also called roll-outs, and to updates using the complete roll-out. 
We call these approaches episodic. 
When we do this, we sum up the immediate updates, which we can rearrange to get
\begin{equation}
  \sum_{t=0}^h \nabla_\theta J(\theta) = \mathbb{E}_{\pi_\theta}\left[\sum_{t=0}^h \nabla_\theta{\log\pi_\theta(a|s)}\psi(\tau)\right].
  \label{eqn:egrad}
\end{equation}

Episodic REINFORCE \cite{Williams92simplestatistical}, the first approach to this concept performs roll-outs and takes only the final return:
\begin{equation}
  \nabla_\theta J(\theta) \approx \frac{1}{h} \sum_{t=0}^h \nabla_\theta\log\pi_\theta(a|s) \mathcal{R}_t^h.
\end{equation}
This is sometimes also called Monte-Carlo gradient estimation. 
However, taking this a step further, it should be possible to attribute the reward at each step to the respective action. 
A very basic episodic score function that fits into our framework in \eqref{eqn:egrad} is to use the discounted rewards, similar to \eqref{eqn:acc-reward}:
\begin{equation}
  \psi_t \leftarrow \gamma^t r_t.
\end{equation}
If we only take batches of episodes starting at time $t'$, we can also use only the rewards following a chosen action $a_{t'}$ by calculating only $\sum_{t=t'}^h \nabla_\theta J(\theta)$ with $\psi_t \leftarrow \gamma^{t'} r_{t}$. 
As we will see, this concept is also used in value functions \eqref{eqn:v}, and \eqref{eqn:q}.

@@

Approximating this function by an unbiased estimator $f_w(s_t, a_t) = \hat{Q}^\pi(s_t, a_t)$, we can also use this estimation as a score function $\psi \leftarrow \hat{Q}$. Sutton et al. \cite{Sutton:1999:PGM:3009657.3009806} show that using this function approximation we will converge to the same local optimum.
However, they also find this requires $\int_\mathbb{S}{\pi(s,a)f_w(s,a)da} = 0$ for each state. 
Given this assumption, the function estimator $f_w$ resembles an advantage function
\begin{equation}
	A^{\pi}(s_t, a_t) := Q^{\pi}(s_t, a_t) - V^{\pi}(s_t)
	\label{eqn:adv}
\end{equation}
as in \cite{1993b}.

>>>>

\paragraph{Definitions.} 
We define states $s \in \mathbb{S}$, actions $a \in \mathbb{A}$, and rewards $r \in \mathbb{R}$. 
For discrete problems, we assume $\mathbb{S} \equiv \{0,1,2,\ldots\}$ and $\mathbb{A} \equiv \{0,1,2,\ldots\}$ without loss of generality.
A trajectory $\tau_{\pi_\theta}^h := (s_0, a_0, $ $s_1, a_1, \dots, s_h, a_h)$ is generated by drawing $s_0 \sim p(x_0)$ according to the distribution over initial states $p(s_0)$, and successively sampling $a_t \sim \pi_\theta(a_t|s_t)$ according to the policy $\pi_\theta$, and $s_{t+1} \sim p(s_{t+1}|s_t,a_t)$ until the horizon $h$ is reached. 
At each time step, we receive a reward according to $r_t = r(s_t, a_t, s_{t+1})$. 
A trajectory can also be called \textit{roll-out} or \textit{episode}, though the term episode implies it ends in a terminal state.

@@

@@

\paragraph{Likelihood-ratio gradients.} 
For this derivation, we will change the perspective a bit requiring some additional definitions. 
We define ${\mu_\pi}_i = \sum_{t=0}^{\infty} p(s_t=s_i | s_0, \pi)$ \cite{puterman2014markov} as the state distribution, though it does not sum up to one without normalization. Note that $\mu_\pi$ is equivalent to the discounted state visit count $d^\pi$ introduced by Sutton et al. \cite{Sutton:1999:PGM:3009657.3009806}.
We define $P_\pi$ as the transition matrix, i.e. ${P_\pi}_{i,j}=p(s_j|s_i,\pi)$.
$r_\pi$ contains the mean rewards for all states given by ${r_\pi}_i = \int_\mathbb{A} r(s_i,a)\pi_\theta(a|s_i) da$. 
$\mu_0$ is the initial state distribution.
From which it follows that $V_\pi = \mu_\pi r_\pi$
From this, we derive
\begin{align}
	\mu_\pi &= \mu_0 + P_\pi^T \mu_\pi \nonumber \\
	(I-P^T)\mu_\pi &= \mu_0 \nonumber \\
	\mu_\pi^T &= \mu_0^T (I-P_\pi)^{-1} ,
\end{align}
so we can reformulate the problem as 
\begin{align}
	\underset{\theta}{\text{max}}\ J(\theta) &= \mu_0^T r_\pi \\
	\text{s.t.}\ r_\pi &= \gamma P_\pi V_\pi . \nonumber
\end{align}
From which we can derive 
\begin{align*}
	\nabla_\theta J(\theta) &= \nabla_\theta \mu_0^T r_\pi
	\\ &= \mu_0^T \nabla_\theta \gamma P_\pi V_\pi
	\\ &= \mu_0^T (I - \gamma P_\pi)^{-1} (\nabla_\theta r_\pi + \gamma (\nabla_\theta P_\pi) V_\pi)
	\\ &= \mu_\pi^T (\nabla_\theta r_\pi + \gamma (\nabla_\theta P_\pi) r_\pi)
	\\ 
	\\ &= \sum \mu_\theta(s) \nabla_\theta\pi(a|s) Q(s,a)
	\\ &= \sum \mu_\theta(s) \pi(a|s) \nabla_\theta\log\pi(a|s) Q(s,a).
\end{align*}
Where we used $\nabla_\theta \pi_\theta(a|s) = \pi_\theta(a|s)\nabla_\theta\log\pi_\theta(a|s)$, obtained from the likelihood ratio, which gives us the likelihood-ratio gradient,
\begin{equation}
	\nabla_\theta J(\theta) = \mathbb{E}_{\pi_\theta}\left[\nabla_\theta{\log\pi_\theta(a|s)}Q_\pi(s,a)\right].
	\label{eqn:like-grad}
\end{equation}
intuitively meaning we should to increase the probability of actions that lead to higher Q-values.

Obviously, we do not have the true $Q_\pi(s,a)$, thus we need to approximate it by $\hat{Q}_\pi(s,a)$. 
We draw $a \sim \pi(a|s)$, starting in state $s_0 \sim p(s_0)$ and match a function estimator to our observations.
Approximating $Q_\pi(s,a)$ by an unbiased estimator $f_w(s_t, a_t) \equiv \hat{Q}_\pi(s_t, a_t)$, Sutton et al. \cite{Sutton:1999:PGM:3009657.3009806} show that using this function approximation we will converge to the same local optimum.
However, they also find this requires $\int_\mathbb{S}{\pi(s,a)f_w(s,a)da} = 0$ for each state. 
Given this assumption, the function estimator $f_w$ is limited to an advantage function
\begin{equation}
	A^{\pi}(s_t, a_t) := Q^{\pi}(s_t, a_t) - V^{\pi}(s_t)
	\label{eqn:adv}
\end{equation}
as in \cite{1993b}, which is mean-zero. We present the effect of this in greater detail in \ref{sec:ac}, but it does not critically affect the following findings.

\paragraph{Immediate updates.} 
The first class of algorithms developed to update a policy directly were REINFORCE \cite{Williams92simplestatistical} algorithms.
REINFORCE uses immediate updates with a baseline $b$ in its gradient steps:
\begin{equation}
  \psi_t \leftarrow r_t-b.
\end{equation}

Obviously, this does not involve temporal information, and is very biased.
To reduce the estimate's bias, we can do updates from completed roll-outs.

\paragraph{Episodic updates.} 
A common way of reducing the bias in value estimation is to sample trajectories, also called roll-outs, and to updates using the complete roll-out. 
We call these approaches episodic. 
When we do this, we sum up the immediate updates, which we can rearrange to get
\begin{equation}
  \sum_{t=0}^h \nabla_\theta J(\theta) = \mathbb{E}_{\pi_\theta}\left[\sum_{t=0}^h \nabla_\theta{\log\pi_\theta(a|s)}Q_\pi(s,a)\right].
  \label{eqn:egrad}
\end{equation}

Episodic REINFORCE \cite{Williams92simplestatistical}, the first approach to this concept performs roll-outs and takes only the final return:
\begin{equation}
  \nabla_\theta J(\theta) \approx \frac{1}{h} \sum_{t=0}^h \nabla_\theta\log\pi_\theta(s) Q(s_t,a_t).
\end{equation}
This is sometimes also called Monte-Carlo gradient estimation. 
However, taking this a step further, it should be possible to attribute the reward at each step to the respective action.
